---
title: "ST237 examples"
author: "Andrew Simpkin"
date: "26/6/2018"
output: 
      learnr::tutorial:
          progressive: false
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(grader)
knitr::opts_chunk$set(echo = TRUE)
tutorial_options(exercise.checker = grade_learnr)
```

## Binomial 
For a general sequence of $n$ Bernoulli trials we can count the number of successes, $X$. The probability of any *particular* pattern of $x$ successes and $(n-x)$ failures is $$ p^x(1-p)^{n-x}$$ 

The number of arrangements of $x$ successes in $n$ trials $=\displaystyle\binom{n}{x}$

$$P(X=x)=\binom{n}{x}p^x(1-p)^{n-x},\qquad x=0,1,2,\ldots,n$$

This is known as the **binomial distribution**  with parameters $n$ (sample size) and $p$ (success probability).

- Mean $= E(X)=np$
- Variance = $Var(X)=np(1-p) = npq$

## Binomial example
- A university team has 11 games in a season
- Probability of winning each game in $0.4$
- Assume independence of games (is this reasonable?)

Using the dbinom function:
1. Find the probability that the team wins 6 games
2. Find the probability that the team has a winning season, i.e. **wins at least 6 games**.

```{r binom1, exercise=TRUE, exercise.lines = 5}

```

```{r binom1-solution}
dbinom(6,11,0.4)
```

```{r binom1-hint}
dbinom(6,11,0.4)
```

```{r binom1-check}
strict_check(success = "Success!")
```

```{r binom2, exercise=TRUE, exercise.lines = 5}

```

```{r binom2-solution}
sum(dbinom(6:11,11,0.4))
```

```{r binom2-hint}
sum(dbinom(6:11,11,0.4))
```

```{r binom2-check}
strict_check(success = "Success!")
```

## Poisson
The random variable $X$ has a Poisson distribution with parameter $\lambda$ if

$$P(X=x)=f_X(x)=\displaystyle\frac{e^{-\lambda}\lambda^x}{x!},\quad x=0,1,2,\ldots$$

- Mean $= E(X)=\lambda$
- Variance = $Var(X)=\lambda$

## Poisson example
- The number of cars per hour entering McDonald's drive-thru is Poisson distributed with mean
$\lambda=60$.
- Arrivals are **independent**
- $T=$ Total number of cars entering in an hour
$$T\sim  \text{Poisson}(60)$$
Using the dpois function:
1. Find the probability of 70 cars entering McDonald's drive-thru
2. Find the probability of at most 70 cars entering McDonald's drive-thru
3. Find the probability of more than 70 cars entering McDonald's drive-thru

```{r pois1, exercise=TRUE, exercise.lines = 5}

```

```{r pois1-solution}
dpois(70,60)
```

```{r pois1-hint}
dpois(70,60)
```

```{r pois1-check}
strict_check(success = "Success!")
```

```{r pois2, exercise=TRUE, exercise.lines = 5}

```

```{r pois2-solution}
sum(dpois(1:70,60))
```

```{r pois2-hint}
sum(dpois(1:70,60))
```

```{r pois2-check}
strict_check(success = "Success!")
```

```{r pois3, exercise=TRUE, exercise.lines = 5}

```

```{r pois3-solution}
1 - sum(dpois(1:70,60))
```

```{r pois3-hint}
1 - sum(dpois(1:70,60))
```

```{r pois3-check}
strict_check(success = "Success!")
```


