---
title: "Tibble and Readr"
output: 
      learnr::tutorial:
          progressive: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(grader)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.checker = grade_learnr)

bechdel <- read_csv("data/BechdelTest.csv")
```
# Topic 4 - Tibble and Readr

## Tibble

### Creating a tibble

```{r echo=T}
tibble( x = 1:5,
        y = 10,
        z = x^3 - y)
```

Create your own tibble
```{r tibble, exercise=TRUE, exercise.lines = 5}
tibble(...)
```

```{r tibble-hint}
"Add paramaters to tibble: column_name = variable"
```

### Tribble
Creating a tibble using tribble allows individual observations to be written one at a time.
```{r echo = T}
tribble(
  ~id, ~name, ~age,
  123, "Leon", 18,
  456, "Elaine", 21,
  789, "Corey", 20
)
```

Create your own tribble
```{r tribble, exercise=TRUE, exercise.lines = 5}

```

```{r tribble-hint}
tribble(...)
```

### Subsetting
Each column of a tibble can be selected using the $ operator or double square brackets [[]]
```{r echo = T}
t <- tibble(x = 5, y = 10)

t$x

t[["y"]]

t %>% .$y

t %>% .[["x"]]
```

Subset your own tibble
```{r subset_tibble, exercise=TRUE, exercise.lines = 5}

```

```{r subset_tibble-hint}
"Use $ or [[]] to subset a tibble."
```

### Convert to data.frame
```{r echo = T}
t <- tibble(x = 5, y = 10)

as.data.frame(t)
```

## Readr

### Reading csv
Using read_csv
```{r echo = T}
pokemon <- read_csv("data/Pokemon.csv")
class(pokemon)
pokemon
```

### Parsing
Parsing string to logical entries.
str shows the structure as logi
```{r echo = T}
str(parse_logical(c("TRUE", "FALSE", "TRUE")))
```

Parse the following vector of strings, v, as integers. Show the structure using str.
```{r parse, exercise=TRUE, exercise.lines = 5}
v <- c("1", "2", "3")

```

```{r parse-solution}
v <- c("1", "2", "3")
str(parse_integer(v))
```

```{r parse-hint}
str(parse_integer(v))
```

```{r parse-check}
strict_check(success = "integers are parsed.")
```

### Date and Time
```{r echo = T}
parse_date("2018-05-25")
```

```{r echo = T}
parse_date("25/05/2018", "%d/%m/%y")
```

Parse the date "01/02/15" as month, day, year.
```{r date, exercise=TRUE, exercise.lines = 5}

```

```{r date-solution}
parse_date("01/02/15", "%m/%d/%y")
```

```{r date-hint}
parse_date("01/02/15", "%m/%d/%y")
```

```{r date-check}
strict_check(success = "Nice date!")
```

### Writing to File
```{r echo = T}
write_csv(bechdel, "data/bechdel2.csv")
```

Write bechdel to R's custom binary format called RDS as "behcel.rds"
```{r rds, exercise=TRUE, exercise.lines = 5}

```

```{r rds-solution}
write_rds(bechdel,"behcel.rds")
```

```{r rds-hint}
write_rds(bechdel,"behcel.rds")
```

```{r rds-check}
strict_check(success = "Bechdel is written as a RDS")
```


