---
title: "Data Transformation"
output: 
      learnr::tutorial:
          progressive: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(grader)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.checker = grade_learnr)

fifa <- read_csv("data/FifaPlayers.csv")
```

# Topic 2 - Data Transformation

## Objects

Create new objects with <-
```{r object, echo = T}
x <- 1 + 2
```
Objects names must start with a letter and may contain letters, numbers and underscores: _

### Comparisons
Check is 2 + 3 equal to 5
```{r comparison, exercise=TRUE, exercise.lines = 5}
... + ... == ...
```

```{r comparison-solution}
2 + 3 == 5
```

```{r comparison-hint}
2 + 3 == 5
```

```{r comparison-check}
strict_check(success = "It is...")
```

## dplyr
5 key dpylr function:

* filter
* arrange
* select
* mutate
* summarize

### Filter
Filter players who have nationality: "Republic of Ireland" and play in the league: "English Premier League".

```{r filter, exercise=TRUE, exercise.lines = 5}
filter(fifa, nationality == "...", ... == "English Premier League")
```

```{r filter-solution}
filter(fifa, nationality == "Republic of Ireland", league == "English Premier League")
```

```{r filter-hint}
filter(fifa, nationality == "Republic of Ireland", league == "English Premier League")
```

```{r filter-check}
strict_check(success = "Nice filter!")
```

### Logical Operators
Filter players who have nationality: "Republic of Ireland" or have nationality: "Wales"

```{r logical1, exercise=TRUE, exercise.lines = 5}
filter(fifa, ... | ...)
```

```{r logical1-solution}
filter(fifa, nationality == "Republic of Ireland" | nationality == "Wales")
```

```{r logical1-hint}
filter(fifa, nationality == "Republic of Ireland" | nationality == "Wales")
```

```{r logical1-check}
strict_check(success = "Nice filter!")
```

### More filter

Filter players who play for a Machester club, Manchester City or Manchester United, using the %in% operator.

```{r logical2, exercise=TRUE, exercise.lines = 5}
filter(fifa, club %in% c(...,...))
```

```{r logical2-solution}
filter(fifa, club %in% c("Manchester United", "Manchester City"))
```

```{r logical2-hint}
filter(fifa, club %in% c("Manchester United", "Manchester City"))
```

```{r logical2-check}
strict_check(success = "Nice filter!")
```

### Missing Values
NA - not available
```{r}
NA
```
The value is not known or is non applicable

filter only includes results which are TRUE, and excludes results which are FALSE and NA
```{r, echo = T}
t <- tibble(x = c(1,20,NA,10,4,NA))
filter(t, x %in% c(1:10))
```

### Arrange
Arrange the "England" players in descending order of age.
Use desc(age)
```{r arrange, exercise=TRUE, exercise.lines = 5}
arrange(filter(fifa,nationality == "England"),...)
```

```{r arrange-solution}
arrange(filter(fifa,nationality == "England"),desc(age))
```

```{r arrange-hint}
desc(age)
```

```{r arrange-check}
strict_check(success = "Nice arranging!")
```

### Select
Select the columns: overall, pac, sho, pas, dri, def, phy
```{r select, exercise=TRUE, exercise.lines = 5}
select(fifa,...)
```

```{r select-solution}
select(fifa, overall, pac, sho, pas, dri, def, phy)
```

```{r select-hint}
select(fifa, overall, pac, sho, pas, dri, def, phy)
```

```{r select-check}
strict_check(success = "Nice column selecting!")
```

### Mutate
Add a new column kick = sho + pas.
Use mutate.
```{r mutate, exercise=TRUE, exercise.lines = 5}
fifa_sample <- select(fifa,sho,pas)
mutate(fifa_sample,...)
```

```{r mutate-solution}
fifa_sample <- select(fifa,sho,pas)
mutate(fifa_sample,kick = sho + pas)
```

```{r mutate-hint}
mutate(fifa_sample,kick = sho + pas)
```

```{r mutate-check}
strict_check(success = "Nice new column!")
```

### Transmute
Use transmute instead of mutate.
This leaves us with only the new columns.
```{r transmute, exercise=TRUE, exercise.lines = 5}
fifa_sample <- select(fifa,sho,pas)
mutate(fifa_sample,kick=sho+pas)
```

```{r transmute-solution}
fifa_sample <- select(fifa,pac,sho,pas,dri,def,phy)
transmute(fifa_sample,kick=sho+pas)
```

```{r transmute-hint}
transmute(fifa_sample,kick=sho+pas)
```

```{r transmute-check}
strict_check(success = "Nice new column!")
```

### Grouping
Group by nationality and prefers_gk.
Summarize with avg_age = mean(age).
```{r group, exercise=TRUE, exercise.lines = 3}
fifa_small <- group_by(fifa, ...)
summarize(fifa_small,...)
```

```{r group-solution}
fifa_small <- group_by(fifa, nationality, prefers_gk)
summarize(fifa_small, avg_age = mean(age))
```

```{r group-hint}
fifa_small <- group_by(fifa, nationality, prefers_gk)
summarize(fifa_small, avg_age = mean(age))
```

```{r group-check}
strict_check(success = "Nice groupings!")
```

### Pipes
dpylr pipes make it easier to nest functions together into one statement.

Without dpylr pipes, using nested functions:
```{r echo = T}
select(filter(fifa,nationality =="Spain"),name,overall)
```

Without dpylr pipes, using variables:
```{r echo = T}
fifa_sample <- filter(fifa,nationality =="Spain")
select(fifa_sample,name,overall)
```

With dpylr pipes:
```{r echo = T}
fifa %>%
  filter(nationality == "Spain") %>%
  select(name,overall)
```

Replace the code below with dpylr pipes: %>%
```{r pipe, exercise=TRUE, exercise.lines = 3}
fifa_sample1 <- group_by(fifa,nationality)
fifa_sample2 <- summarize(fifa_sample1,count = n(), avg_age = mean(age))
fifa_sample3 <- filter(fifa_sample2,avg_age < 25)
arrange(fifa_sample3, avg_age)
```

```{r pipe-solution}
fifa %>% 
  group_by(nationality) %>%
  summarize(count = n(), avg_age = mean(age)) %>%
  filter(avg_age < 25) %>%
  arrange(avg_age)
```

```{r pipe-hint}
fifa %>% 
  group_by(nationality) %>%
  summarize(count = n(), avg_age = mean(age)) %>%
  filter(avg_age < 25) %>% 
  arrange(avg_age)
```

```{r pipe-check}
strict_check(success = "Nice dplyr pipes!")
```
