---
title: "Exploratory Data Analysis"
output: 
      learnr::tutorial:
          progressive: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(grader)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.checker = grade_learnr)

pokemon <- read_csv("data/Pokemon.csv")

bechdel <- read_csv("data/BechdelTest.csv")

fifa <- read_csv("data/FifaPlayers.csv")
```
# Topic 3 - Exploratory Data Analysis

## Visualizing Distributions
Bar chart showing the number of Pokemon of each type
```{r echo=T}
ggplot(data=pokemon) +
  geom_bar(mapping = aes(x = type_1)) +
  coord_flip()
```

Visualise the distrubtion of pokemon's hp using a histogram with binwidth = 10
```{r visual1, exercise=TRUE, exercise.lines = 5}
ggplot(data=pokemon) +
  ...
```

```{r visual1-solution}
ggplot(data=pokemon) +
  geom_histogram(mapping = aes(x = hp),binwidth=10)
```

```{r visual1-hint}
ggplot(data=pokemon) +
  geom_histogram(mapping = aes(x = hp),binwidth=10)
```

```{r visual1-check}
strict_check(success = "Nice histogram!")
```

Visualise the distrubtion of pokemon's attack using a freqpoly, filtering dragon, ground, normal and water type_1 pokemons, using a freqpoly with color = type_1 and binwidth=5
```{r visual2, exercise=TRUE, exercise.lines = 5}
ggplot(data=filter(pokemon,type_1 %in% ...)) +
```

```{r visual2-solution}
ggplot(data=filter(pokemon,type_1 %in% c("Water","Normal","Grass","Bug"))) +
  geom_freqpoly(mapping = aes(x = attack,color = type_1), binwidth=5)
```

```{r visual2-hint}
ggplot(data=filter(pokemon,type_1 %in% c("Water","Normal","Grass","Bug"))) +
  geom_freqpoly(mapping = aes(x = attack,color = type_1), binwidth=5)
```

```{r visual2-check}
strict_check(success = "Nice histogram!")
```

## Unusual Values
Histogram of wage of footballers
```{r echo=T}
ggplot(data=fifa) +
  geom_histogram(mapping = aes(x = eur_wage),binwidth=1000)
```
The only evidence of outliers is the unusually wide limits on the y-axis

Add coord_cartesian ylim of c(0,20)
```{r unusual, exercise=TRUE, exercise.lines = 5}
ggplot(data=fifa) +
  geom_histogram(mapping = aes(x = eur_wage),binwidth=1000)
```

```{r unusual-solution}
ggplot(data=fifa) +
  geom_histogram(mapping = aes(x = eur_wage),binwidth=1000) +
  coord_cartesian(ylim = c(0,20))
```

```{r unusual-hint}
ggplot(data=fifa) +
  geom_histogram(mapping = aes(x = eur_wage),binwidth=1000) +
  coord_cartesian(ylim = c(0,20))
```

```{r unusual-check}
strict_check(success = "Well done!. It is much easier to see small values now.")
```

## Missing Values
NA values can not be plotted and may give errors
```{r echo=T}
ggplot(data=bechdel,mapping = aes(x = test, y = dom_gross)) +
  geom_point()
```
Remove the missing value warning by using the paramater na.rm = TRUE in geom_point

```{r missing, exercise=TRUE, exercise.lines = 5}
ggplot(data=bechdel,mapping = aes(x = test, y = dom_gross)) +
  geom_point()
```

```{r missing-solution}
ggplot(data=bechdel,mapping = aes(x = test, y = dom_gross)) +
  geom_point(na.rm = TRUE)
```

```{r missing-hint}
ggplot(data=bechdel,mapping = aes(x = test, y = dom_gross)) +
  geom_point(na.rm = TRUE)
```

```{r missing-check}
strict_check(success = "Missing values no longer give warnings.")
```

## Covariation
Covariation is the tendency for the values of two or more variables to vary togeter in a related way.

International Reputation vs Average Wage
```{r echo=T}
ggplot(data=(fifa %>% 
  group_by(international_reputation) %>%
  summarize(count = n(), avg_wage = mean(eur_wage))))+
  geom_point(mapping = aes(x = international_reputation, y = avg_wage))
```

Find covariation between two other variables in the fifa data
```{r covariation, exercise=TRUE, exercise.lines = 5}
ggplot(data=fifa,mapping = aes(...)) +
  ...
```

```{r covariation-hint}
"Plot two variables in the fifa dataset that may be related."
```

## Patterns
From this tile plot we can see a decline in number of water pokemon through generations, and a reduced number of pokemon in generation 6 when compared to other generations.

```{r echo=T}
pokemon %>%
  count(generation,type_1) %>%
  ggplot(mapping = aes(x=generation,y=type_1)) +
    geom_tile(mapping = aes(fill = n))
```

This barchart confirms the suspician that the 6th generation of Pokemon has the least number of Pokemon.
```{r echo=T}
  ggplot(data=pokemon,mapping = aes(x = generation)) +
  geom_bar()
```
