---
title: "Data Visualisation with ggplot2"
output: 
      learnr::tutorial:
          progressive: false
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(grader)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.checker = grade_learnr)

pokemon <- read_csv("data/Pokemon.csv")

bechdel <- read_csv("data/BechdelTest.csv")

fifa <- read_csv("data/FifaPlayers.csv")
```


# Topic 1 - Data Visualisation with ggplot2

## Scatter Plots
Pokemon data of all Pokemon (generations 1 to 6)
```{r echo = T}
pokemon
```

Here is a simple ggplot of pokemon defense vs attack
```{r ggplot_simple}
ggplot(data = pokemon) +
  geom_point(mapping = aes(x = attack, y = defense))
```

### A graphing workflow
Our code follows the common workflow for making graphs with ggplot2. To make a graph, you:

Start the graph with ggplot()
Add elements to the graph with a geom_ function
Select variables with the mapping = aes() argument

ggplot template:
 ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))

### Color
Color the points by the type of each pokemon by inserting color = generation to the paramaters in aes
```{r ggplot_color, exercise=TRUE, exercise.lines = 5}
ggplot(data = pokemon) +
  geom_point(mapping = aes(x = attack, y = defense))
```

```{r ggplot_color-solution}
ggplot(data = pokemon) +
  geom_point(mapping = aes(color = generation, x = attack, y = defense))
```

```{r ggplot_color-hint}
aes(x = attack, y = defense, color = generation)
```

```{r ggplot_color-check}
strict_check(success = "Nice colors!")
```
### Shape
Shape the points by the generation of each pokemon by inserting shape = generation to the paramaters in aes. To assign shapes to a continuous variable add + scale_shape_identity() to the ggplot

```{r ggplot_shape, exercise=TRUE, exercise.lines = 5}
ggplot(data = pokemon) +
  geom_point(mapping = aes(x = attack, y = defense))
```

```{r ggplot_shape-solution}
ggplot(data = pokemon) +
  geom_point(mapping = aes(shape = generation, x = attack, y = defense)) +
    scale_shape_identity()
```

```{r ggplot_shape-hint}
aes(x = attack, y = defense, shape = generation) +
    scale_shape_identity()
```

```{r ggplot_shape-check}
strict_check(success = "Nice shapes! ggplot2 will only use six shapes at a time, additional groups will go unplotted when you use this aesthitc.")
```

### Facets
Add + facet_wrap(~ generation, nrow = 2) to the ggplot, this facets the plot on the generation. Set the color = legendary

```{r ggplot_facet, exercise=TRUE, exercise.lines = 5}
ggplot(data = pokemon) +
  geom_point(mapping = aes(x = attack, y = defense))
```

```{r ggplot_facet-solution}
ggplot(data = pokemon) +
  geom_point(mapping = aes(x = attack, y = defense, color = legendary)) +
  facet_wrap(~ generation, nrow = 2)
```

```{r ggplot_facet-hint}
 geom_point(mapping = aes(x = attack, y = defense, color = legendary)) +
  facet_wrap(~ generation, nrow = 2)
```

```{r ggplot_facet-check}
strict_check(success = "Nice facet! Notice generation 2 and 3 has pokemon with the highest defense and the plot of generation 5 does not have very extreme outliers compared to the other generations.")
```

### Smooth Geom
Scatter plot of Pokemons' defense vs hp
```{r}
ggplot(data = pokemon) +
  geom_point(mapping = aes(x = hp, y = defense))
```

Change geom_point to geom_smooth with paramter methos = 'loess'
```{r ggplot_smooth, exercise=TRUE, exercise.lines = 5}
ggplot(data = pokemon) +
  geom_point(mapping = aes(x = hp, y = defense))
```

```{r ggplot_smooth-solution}
ggplot(data = pokemon) +
  geom_smooth(method = 'loess', mapping = aes(x = hp, y = defense))
```

```{r ggplot_smooth-hint}
geom_smooth(method = 'loess', mapping = aes(x = hp, y = defense))
```

```{r ggplot_smooth-check}
strict_check(success = "Nice smooth curve!")
```

### Overlay Graphs
How can we overly the two previous graphics on defense vs hp?

Add the geom_smooth to the ggplot using the + operator
```{r ggplot_overlay, exercise=TRUE, exercise.lines = 5}
ggplot(data = pokemon) +
  geom_point(mapping = aes(x = hp, y = defense))
```

```{r ggplot_overlay-solution}
ggplot(data = pokemon) +
  geom_smooth(method = 'loess',mapping = aes(x = hp, y = defense)) +
  geom_point(mapping = aes(x = hp, y = defense))
```

```{r ggplot_overlay-hint}
  geom_smooth(method = 'loess',mapping = aes(x = hp, y = defense)) +
  geom_point(mapping = aes(x = hp, y = defense))
```

```{r ggplot_overlay-check}
strict_check(success = "Nice overlay!")
```

## Bar Charts
The Bechdel Test tests the representation of women in films.
The test asks 3 questions in order:
1. Are there women?
2. Do the women talk to each other?
3. Do the women talk about something other than men?
```{r echo = T}
bechdel
```

Bar chart of number of films that passed and failed the bechdel test
```{r}
ggplot(data = bechdel) +
  geom_bar(mapping = aes(x = test))
```

### Fill
Bars are filled by proportion of how many bechdel test questions the films passed.
```{r}
ggplot(data = bechdel) +
  geom_bar(mapping = aes(x = binary, fill = test))
```

### Fill Position
Add position = "fill" in geom_bar
```{r ggplot_fill, exercise=TRUE, exercise.lines = 5}
ggplot(data = bechdel) +
  geom_bar(mapping = aes(x = binary, fill = test))
```

```{r ggplot_fill-solution}
ggplot(data = bechdel) +
  geom_bar(mapping = aes(x = binary, fill = test), position = "fill")
```

```{r ggplot_fill-hint}
  geom_bar(mapping = aes(x = binary, fill = test), position = "fill")
```

```{r ggplot_fill-check}
strict_check(success = "Nice filled bar chart! We can see that most films that do not pass the test, have women in the film that do not talk to each other.")
```

### Dodge Position
Add position = "dodge" to geom_bar
```{r ggplot_dodge, exercise=TRUE, exercise.lines = 5}
ggplot(data = bechdel) +
  geom_bar(mapping = aes(x = binary, fill = test))
```

```{r ggplot_dodge-solution}
ggplot(data = bechdel) +
  geom_bar(mapping = aes(x = binary, fill = test), position = "dodge")
```

```{r ggplot_dodge-hint}
  geom_bar(mapping = aes(x = binary, fill = test), position = "dodge")
```

```{r ggplot_dodge-check}
strict_check(success = "Nice filled bar chart with dodge!")
```

## Box Plots
Fifa 18 player data
```{r echo = T}
fifa
```
### Top 5 clubs in EPL
Add + geom_boxplot() to the ggplot
```{r ggplot_box, exercise=TRUE, exercise.lines = 5}
ggplot(data = filter(fifa[1:100,],league =="English Premier League"), mapping = aes(x = club, y = overall)) + 
  labs(y = "overall rating")

```

```{r ggplot_box-solution}
ggplot(data = filter(fifa[1:100,],league =="English Premier League"), mapping = aes(x = club, y = overall)) + 
  labs(y = "overall rating") + 
  geom_boxplot()
```

```{r ggplot_box-hint}
  + geom_boxplot()
```

```{r ggplot_box-check}
strict_check(success = "Nice boxplot!")
```

### Coorinate System.
Add + coord_flip() to the ggplot
```{r ggplot_box_flip, exercise=TRUE, exercise.lines = 5}
ggplot(data = filter(fifa,league =="English Premier League"), mapping = aes(x = club, y = overall)) + 
  labs(y = "overall rating")+ 
  geom_boxplot()
```

```{r ggplot_box_flip-solution}
ggplot(data = filter(fifa,league =="English Premier League"), mapping = aes(x = club, y = overall)) + 
  labs(y = "overall rating") + 
  geom_boxplot()+ 
  coord_flip()
```

```{r ggplot_box_flip-hint}
  + coord_flip()
```

```{r ggplot_box_flip-check}
strict_check(success = "Nice flipped boxplot! Now all the team names are visible.")
```