---
title: "Relational Data"
output: 
      learnr::tutorial:
          progressive: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(grader)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.checker = grade_learnr)

pokemon_combat <- read_csv("data/PokemonCombat.csv")
pokemon_type <- read_csv("data/PokemonType.csv")

pokemon_combat_g1 <- filter(pokemon_combat,generation == 1)
```
# Topic 6 - Relational Data

Multiple tables of data are called relational data. It is realtions, not just the individual datasets that are important. 
Mutate Joins: add new variables to one data frame from mathching observations in another.

Filtering joins: filter observations from one data frame based on whether they match an observation in the other table.

Set operations: treat observations as if they were set elements.

Two related data frames.
One describes the Pokemon's combat ability.
The other describes the Pokemons' type
```{r echo = T}
pokemon_combat

pokemon_type
```

## Inner Join

inner_join joins two tables by a common key.
```{r echo = T}
pokemon_combat_g1 %>%
  inner_join(pokemon_type, by = "#")
```

## Outer Join
Outer join joins tables by a key and keeps observations that appear in the left or right table.  

left_join of pokemon_combat_g1 with pokemon_type by "#"
```{r echo = T}
pokemon_combat_g1 %>%
  left_join(pokemon_type, by = "#")
```

right_join pokemon_combat with pokemon_type by "#"
```{r outer, exercise=TRUE, exercise.lines = 5}
pokemon_combat_g1 %>%
```

```{r outer-solution}
pokemon_combat_g1 %>%
  right_join(pokemon_type, by = "#")
```

```{r outer-hint}
pokemon_combat_g1 %>%
  right_join(pokemon_type, by = "#")
```

```{r outer-check}
strict_check(success = "Nice outer join!")
```

## Filter Join
semi_join pokemon_combat_g1 with pokemon_type by "#"
```{r filter, exercise=TRUE, exercise.lines = 5}
pokemon_combat_g1 %>%
```

```{r filter-solution}
pokemon_combat_g1 %>%
  semi_join(pokemon_type, by = "#")
```

```{r filter-hint}
pokemon_combat_g1 %>%
  semi_join(pokemon_type, by = "#")
```

```{r filter-check}
strict_check(success = "Nice semi join!")
```




